The Effects of weather on maize yield: New evidence from Kenya

### 2019-01-15 writing/draft4/draft23.tex
writing MAPE description

\address{Department of Geography, University of Sussex, Falmer, UK}
\ead{monika.novac@gmail.com}
\vspace{10pt}
\begin{indented}
\item[]December 2018
\end{indented}

@@

Following a standard practise, we calculated the variance inflation factors (VIF) of the explanatory variables to test for presence of multicollinearity. With only two exceptions, the VIF was below $2$ for all predictors in our final models. For CV of precipitation, the value of VIF was $2.172$ in the model of all counties, $2.147$ in the model for the subsample of the ASAL counties and it was equal to $2.182$ in the model based on the subsample of the non-ASAL counties. These values are well below the threshold of problematic multicollinearity. The other exception was the linear term of seasonal precipitation as its VIF was equal to $2.407$ in the model for the non-ASAL counties, probably due to presence of its squared term in the model. However, $2.407$ is below the threshold of the problematic multicollinearity and the VIF of precipitation was below $2$ in the other two models. Therefore, we can conclude that the models do not suffer from presence of multicollinearity.

To evaluate out of sample prediction accuracy of our models we implemented a procedure which is sometimes called the \textit{jackknife}. We re-estimated the preferred mixed effects models $n$-times, where $n$ is the number of observations (in our case $1300$ for the sample of all counties and $698$ and $602$ for the subsamples of ASAL and non-ASAL counties, respectively), each time we left out one observation. We subsequently used each of these $n$ estimated models to generate a predicted value 
of log of maize yield of the left-out observation. Then we calculated the absolute value of the difference between the predicted and real value of the left-out observation for each of the $n$ models. In the next step, we averaged the $n$ absolute values of the differences between predicted and real values to obtain the mean absolute prediction error (MAPE). Finally, we scaled the MAPE values by dividing them by the interquartile range (IQR) of the sample obtained as a difference between the third and first quartiles.

Besides obtaining the MAPE of our preferred specifications, we calculated MAPE of baseline models specified as mixed effects models with the set of fixed effects including two measures, in particular cumulative seasonal precipitation and seasonal average temperature. The only random effects of the baseline models are random intercepts as in our preferred specification and the baseline models have the same error structure ARMA($1$,$1$) as our main models. For the sake of comparison, we further calculated MAPE of naive forecasts defined as the last period's actual values of each county.

The scaled MAPE of our preferred specification and the two baseline predictions can be found in table~\ref{CVdiv}. All three predictions were obtained for the sample of all counties and also for the models based on the subsamples of ASAL and non-ASAL counties. The values of MAPE not divided by IQR can be found in table~\ref{CV} in the appendix.

@@

According to MAPE, our preferred specifications perform better than the baseline mixed effects models and considerably better than the naive forecasts for all three samples (table~\ref{CVdiv}). For all three prediction methods, the smallest values of MAPE were achieved for the sample of all counties while the least precise forecasts with regard to MAPE were obtained for the subsample of ASAL counties (table~\ref{CVdiv}).

>>>>

\address{Department of Geography, University of Sussex, Falmer, UK}
\ead{monika.novac@gmail.com}
\vspace{10pt}
\begin{indented}
\item[]January 2019
\end{indented}

@@

Following a standard practise, we calculated the variance inflation factors (VIF) of the explanatory variables to test for presence of multicollinearity. With only two exceptions, the VIF was below $2$ for all predictors in our final models. For CV of precipitation, the value of VIF was $2.172$ in the model of all counties, $2.147$ in the model for the subsample of the ASAL counties and it was equal to $2.182$ in the model based on the subsample of the non-ASAL counties. These values are well below the threshold of problematic multicollinearity. The other exception was the linear term of precipitation as its VIF was equal to $2.407$ in the model for the non-ASAL counties, probably due to presence of its squared term in the model. However, $2.407$ is below the threshold of the problematic multicollinearity and the VIF of precipitation was below $2$ in the other two models. Therefore, we can conclude that the models do not suffer from presence of multicollinearity.

To evaluate out of sample prediction accuracy of our models we implemented a procedure which is sometimes called the \textit{jackknife}. We re-estimated the preferred mixed effects models $n$-times, where $n$ is the number of observations (in our case $1300$ for the sample of all counties and $698$ and $602$ for the subsamples of ASAL and non-ASAL counties, respectively), each time we left out one observation. We subsequently used each of these $n$ estimated models to generate a predicted value 
of log of maize yield of the left-out observation. After that, we determined the absolute value of the difference between the predicted and real value of the left-out observation for each of the $n$ models. In the next step, we calculated average of the $n$ absolute values of the differences between predicted and real values to obtain the mean absolute prediction error (MAPE). Finally, we scaled the MAPE values by dividing them by the interquartile range (IQR) of the sample obtained as a difference between the third and first sample quartiles.

Besides obtaining the MAPE of our preferred specification, we worked out MAPE of a baseline model specified as a mixed effects model with the set of fixed effects including two measures, in particular cumulative seasonal precipitation and seasonal average temperature. The only random effects of the baseline model were random intercepts as in our preferred specification. The baseline model had the error structure ARMA($1$,$1$) as our main model. For the sake of comparison, we further calculated MAPE of naive forecasts defined as the last period's actual values of each county.

The scaled MAPE of our preferred specification and the two baseline predictions can be found in table~\ref{CVdiv}. All three predictions were obtained for the full-sample model of all counties and also for the models based on the subsamples of ASAL and non-ASAL counties. The unscaled values of MAPE (not divided by the IQR) can be found in table~\ref{CV} in the appendix.

@@

According to the MAPE, our preferred specifications perform better than the baseline mixed effects models and they perform considerably better than the naive forecasts for all three samples (table~\ref{CVdiv}). For all three prediction methods, the smallest values of MAPE were achieved for the sample of all counties while the least precise forecasts (with regard to MAPE) were obtained for the subsample of ASAL counties (table~\ref{CVdiv}).

### 2019-01-15 writing/draft4/draftMonika.tex
MAPE asal non asal
writing MAPE description

To verify our results, we adapted an alternative step-down model building
approach using the Satterthwaite’s method to determine the p-values of the individual
t-tests \cite{lmerTest}. We started with a model which included the complete set of our weather measures in both fixed effects and random effects. In the first part of the procedure, the insignificant random effects were removed one by one. In each iteration, the insignificant variable with the highest p-value was removed. This was repeated until only the significant random effects remained. After this, we applied an analogous procedure to the fixed effects until we were left only with the significant variables. As recommended in literature, we considered the level of significance $\alpha=0.1$ for the random effects and $\alpha=0.05$~ for the fixed effects \cite{lmerTest}. This method led to the same model as the one obtained with our primary AIC-based method described above.

According to the conditional Lagrange multiplier (LM) test developed by Baltagi \& Li (1991) and Baltagi \& Li (1995), the errors exhibited a within group autcorrelation structure in our models (the p-value of the LM test was $1.1\times10^{-13}$). To further investigate the autocorrelation structure of the errors, we estimated a number of models (with the subset of predictors chosen as described above) with an ARMA($p$,$q$) error structure. In particular, we estimated all variants of our model with ARMA($p$,$q$) errors such that $p\leq2$ and $q\leq2$. Comparing the AIC criteria and using the corresponding likelihood ratio statistics, we found that the most appropriate error correlation structure was ARMA($1$,$1$). The value of the AIC criterion of the model with the ARMA($1$,$1$) error structure was $2122.2$ while the AIC was $2129.2$ for the model with ARMA($1$,$0$) errors.  The p-value of the likelihood ratio test of the comparison of the model with ARMA($1$,$1$) errors and the model with ARMA($1$,$0$) errors was smaller than $0.003$, hence the ARMA($1$,$1$) error structure turned out to be a better fit. The AIC criteria and the likelihood ratio tests of all the models with ARMA($p$,$q$) errors are summarised in table~\ref{ARMA} in the appendix.

@@

The AIC is an important criterion for comparison of competing models, however it does not provide any information about the absolute goodness of fit. In order to assess the absolute goodness of fit of a model, a different measure is needed. For a simple linear regression, the coeffiecient of determination $R^2$ or its adjusted version is typically used for this purpose. Nevertheless, obtaining the $R^2$ or the adjusted $R^2$ is not straightforward for linear mixed models or for models estimated by the maximum likelihood in general. In this study, we used the coefficient of determination of generalized linear mixed-effects models introduced by \cite{Nakagawa2013}. For each of our samples (all counties, ASAL and non-ASAL), we present two versions of the statistic in table~\ref{R2}. These include marginal $R^2$ which can be interpreted as the variance explained by the fixed effects and the conditional $R^2$ which can be interpreted as the variance explained by both fixed effects and random effects. In our case, the only random effects are the intercepts, hence the difference between the two variants is the variance explained by the random intercepts. As is apparent from table~\ref{R2}, the fixed effects explain about $30\%$ of the variance in case of the model for all counties,  about $13\%$ of the variance in the model based on the ASAL counties and about $15\%$ of the variance in the model based on the non-ASAL counties.

\begin{center}
\begin{threeparttable}
\caption{\label{R2}Coefficient of determination $R^2$ for mixed models}
\begin{indented}
\item[]
\begin{tabular}{@{}lcc}
\\[-1em]
\br
\\[-1em]
\textbf{Sample:}&Marginal\tnote{a}&Conditional\tnote{b}\\
\hline
{All counties}& $0.300$&$0.526$\\
{ASAL}& $0.134$&$0.353$\\
{non-ASAL}&  $0.153$&$0.436$\\
\br
\end{tabular}
\begin{tablenotes}
\begin{footnotesize}
    \item \textit{Notes:} The coefficient of determination from generalised linear mixed-effects models. The statistic is based on \cite{Nakagawa2013}.  \\
        \begin{adjustwidth}{1cm}{}
    \item[a] The marginal $R^2$ is interpreted as the variance explained by the fixed effects
  \item[b] The conditional $R^2$ is explained by both fixed affects and random effects (in our case by fixed effects and random intercepts)
     \end{adjustwidth}
     \end{footnotesize}
\end{tablenotes}

\end{indented}
\end{threeparttable} 
\end{center}
\vspace{1cm}

Following a standard practise, we calculated the variance inflation factors (VIF) of the explanatory variables to test for presence of multicollinearity. With only two exceptions, the VIF is below $2$ for all predictors in our final models. For CV of precipitation, the value of VIF is $2.172$ in the model of all counties, $2.147$ in the model for the subsample of the ASAL counties and it is equal to $2.182$ in the model based on the subsample of the non-ASAL counties. These values are well below the threshold of problematic multicollinearity. The other exception is the linear term of seasonal precipitation as its VIF is $2.407$ in the model for the non-ASAL counties, probably due to presence of the squared term of precipitation. However, $2.407$ is below the threshold of the problematic multicollinearity and the VIF of precipitation is below $2$ in the other two models. Therefore, we can conclude that the models do not suffer from presence of multicollinearity.

%\item Possibly estimate simple regressions with the same explanatory variables as the preferred mixed effects specification. This would give us an approximation of $R^2$ which could be used to compare our models.

@@

\begin{itemize}
\color{blue}
\item If we get the yield data for the period from $2015$ onwards: Out of sample predictions and comparison with the real data
\end{itemize}

\color{black}

\appendix
\section*{Appendix}

@@

\section*{References}

\bibliography{referencesFS}

>>>>

To verify our results, we adapted an alternative step-down model building
approach using the Satterthwaite’s method to determine the p-values of the individual
t-tests \cite{lmerTest}. We started with a model which included the complete set of our weather measures in both fixed effects and random effects. In the first part of the procedure, insignificant random effects were removed one by one. In each iteration, the insignificant variable with the highest p-value was removed. This was repeated until only significant random effects remained. After this, we applied an analogous procedure to the fixed effects until we were left only with significant variables. As recommended in literature, we considered the level of significance $\alpha=0.1$ for random effects and $\alpha=0.05$ for fixed effects \cite{lmerTest}. This method led us to the same results as those obtained with our primary AIC-based method described above.

According to the conditional Lagrange multiplier (LM) test developed by Baltagi \& Li (1991; 1995), the errors exhibited a within group autcorrelation structure in our models (the p-value of the LM test was $1.1\times10^{-13}$). To further investigate the autocorrelation structure of the errors, we estimated a number of models (with the subset of predictors chosen as described above) with an ARMA($p$,$q$) error structure. In particular, we estimated all variants of our model with ARMA($p$,$q$) errors such that $p\leq2$ and $q\leq2$. Comparing the AIC criteria and using the corresponding likelihood ratio test statistics, we found that the most appropriate error correlation structure was ARMA($1$,$1$). The value of the AIC criterion of the model with the ARMA($1$,$1$) error structure was $2122.2$ while the AIC was $2129.2$ for the model with ARMA($1$,$0$) errors.  The p-value of the likelihood ratio test of the comparison of the model with ARMA($1$,$1$) errors and the model with ARMA($1$,$0$) errors was smaller than $0.003$, hence the ARMA($1$,$1$) error structure turned out to be a better fit. The AIC criteria and the likelihood ratio tests of all the models with ARMA($p$,$q$) errors are summarised in table~\ref{ARMA} in the appendix.

@@

The AIC is an important criterion for comparison of competing models, however it does not provide any information about the absolute goodness of fit. In order to assess the absolute goodness of fit of a model, a different measure is needed. For a simple linear regression, the coeffiecient of determination $R^2$ or its adjusted version is typically used for this purpose. Nevertheless, obtaining the $R^2$ or the adjusted $R^2$ is not straightforward for linear mixed models or for models estimated by the maximum likelihood in general. In this study, we used the coefficient of determination of generalized linear mixed-effects models introduced by \cite{Nakagawa2013}. For each of our samples (all counties, ASAL and non-ASAL), we present two versions of the generalised $R^2$ in table~\ref{R2}. These include marginal $R^2$ which can be interpreted as the variance explained by the fixed effects and the conditional $R^2$ which can be interpreted as the variance explained by both fixed effects and random effects. In our case, the only random effects are the intercepts, hence the difference between the two variants is the variance explained by the random intercepts. As is apparent from table~\ref{R2}, the fixed effects explain about $30\%$ of the variance in case of the model for all counties,  about $13\%$ of the variance in the model based on the ASAL counties and about $15\%$ of the variance in the model based on the non-ASAL counties.

\begin{center}
\begin{threeparttable}
\caption{\label{R2}Coefficient of determination $R^2$ for mixed models}
\begin{indented}
\item[]
\begin{tabular}{@{}lcc}
\\[-1em]
\br
\\[-1em]
\textbf{Sample:}&Marginal\tnote{a}&Conditional\tnote{b}\\
\hline
{All counties}& $0.300$&$0.526$\\
{ASAL}& $0.134$&$0.353$\\
{non-ASAL}&  $0.153$&$0.436$\\
\br
\end{tabular}
\begin{tablenotes}
\begin{footnotesize}
    \item \textit{Notes:} The coefficient of determination for generalised linear mixed-effects models \cite{Nakagawa2013}.  \\
        \begin{adjustwidth}{1cm}{}
    \item[a] The marginal $R^2$ is interpreted as the variance explained by the fixed effects
  \item[b] The conditional $R^2$ is explained by both fixed affects and random effects (in our case by fixed effects and random intercepts)
     \end{adjustwidth}
     \end{footnotesize}
\end{tablenotes}     
\end{indented}
\end{threeparttable} 
\end{center}
\vspace{1cm}

@@

Following a standard practise, we calculated the variance inflation factors (VIF) of the explanatory variables to test for presence of multicollinearity. With only two exceptions, the VIF was below $2$ for all predictors in our final models. For CV of precipitation, the value of VIF was $2.172$ in the model of all counties, $2.147$ in the model for the subsample of the ASAL counties and it was equal to $2.182$ in the model based on the subsample of the non-ASAL counties. These values are well below the threshold of problematic multicollinearity. The other exception was the linear term of precipitation as its VIF was equal to $2.407$ in the model for the non-ASAL counties, probably due to presence of its squared term in the model. However, $2.407$ is below the threshold of the problematic multicollinearity and the VIF of precipitation was below $2$ in the other two models. Therefore, we can conclude that the models do not suffer from presence of multicollinearity.

To evaluate out of sample prediction accuracy of our models we implemented a procedure which is sometimes called the \textit{jackknife}. We re-estimated the preferred mixed effects models $n$-times, where $n$ is the number of observations (in our case $1300$ for the sample of all counties and $698$ and $602$ for the subsamples of ASAL and non-ASAL counties, respectively), each time we left out one observation. We subsequently used each of these $n$ estimated models to generate a predicted value 
of log of maize yield of the left-out observation. After that, we determined the absolute value of the difference between the predicted and real value of the left-out observation for each of the $n$ models. In the next step, we calculated average of the $n$ absolute values of the differences between predicted and real values to obtain the mean absolute prediction error (MAPE). Finally, we scaled the MAPE values by dividing them by the interquartile range (IQR) of the sample obtained as a difference between the third and first sample quartiles.

Besides obtaining the MAPE of our preferred specification, we worked out MAPE of a baseline model specified as a mixed effects model with the set of fixed effects including two measures, in particular cumulative seasonal precipitation and seasonal average temperature. The only random effects of the baseline model were random intercepts as in our preferred specification. The baseline model had the error structure ARMA($1$,$1$) as our main model. For the sake of comparison, we further calculated MAPE of naive forecasts defined as the last period's actual values of each county.

The scaled MAPE of our preferred specification and the two baseline predictions can be found in table~\ref{CVdiv}. All three predictions were obtained for the full-sample model of all counties and also for the models based on the subsamples of ASAL and non-ASAL counties. The unscaled values of MAPE (not divided by the IQR) can be found in table~\ref{CV} in the appendix.

\begin{center}
\begin{threeparttable}
\caption{\label{CVdiv}	Leave-one-out cross validation: Mean absolute prediction error divided by IQR}
\linespread{1.3}
% files Rcodes/crossvalidation/compareJustTotals.R; Rcodes/crossvalidation/MAPE_ASAL.R;  Rcodes/crossvalidation/MAPE_nonASAL.R  
\begin{indented}
\item[]
\begin{tabular}{@{}lccc}
\\[-1em]
\br
\\[-1em]
\textbf{Sample}&\textbf{Preferred specification\tnote{a}}&\textbf{Baseline specification\tnote{b}}&\textbf{Naive forecast\tnote{c}}\\
\hline
{All counties}&$0.3811$&$0.3889$&$0.4277$\\
{ASAL}&$0.5300$&$0.5342$&$0.5835$\\
{non-ASAL}&$0.3858$&$0.4043$&$0.4510$\\
\br
\end{tabular}
\end{indented}
 \begin{tablenotes}
  \begin{footnotesize}
  \item[a] Mixed effects models: the preferred specification with all fixed effects listed in table~\ref{MainEst}.
  \item[b] Mixed effects models: the baseline specification, only seasonal precipitation and average seasonal temperature in the fixed effects.
  \item[c] Last period's actual values used as this period's forecast.
  \end{footnotesize}
\end{tablenotes}
\end{threeparttable} 
\end{center}
\vspace{1cm}

According to the MAPE, our preferred specifications perform better than the baseline mixed effects models and they perform considerably better than the naive forecasts for all three samples (table~\ref{CVdiv}). For all three prediction methods, the smallest values of MAPE were achieved for the sample of all counties while the least precise forecasts (with regard to MAPE) were obtained for the subsample of ASAL counties (table~\ref{CVdiv}).

\clearpage
\section{Conclusions}\label{Conclusions}
\clearpage
\appendix
\section*{Appendix}

@@

\begin{center}
\begin{threeparttable}
\caption{\label{CV}	Leave-one-out cross validation: Mean absolute prediction error}
\linespread{1.3}
% files Rcodes/crossvalidation/compareJustTotals.R; Rcodes/crossvalidation/MAPE_ASAL.R;  Rcodes/crossvalidation/MAPE_nonASAL.R  
\begin{indented}
\item[]
\begin{tabular}{@{}lccc}
\\[-1em]
\br
\\[-1em]
\textbf{Sample}&\textbf{Preferred specification\tnote{a}}&\textbf{Baseline specification\tnote{b}}&\textbf{Naive forecast\tnote{c}}\\
\hline
{All counties}&$0.3599$&$0.3673$&$0.4040$\\
{ASAL}&$0.4352$&$0.4385$&$0.4790$\\
{non-ASAL}&$0.2715$&$0.2845$&$0.3173$\\
\br
\end{tabular}
\end{indented}

 \begin{tablenotes}
  \begin{footnotesize}
  \item[a] Mixed effects models: the preferred specification with all fixed effects listed in table~\ref{MainEst}.
  \item[b] Mixed effects models: the baseline specification, only seasonal precipitation and average seasonal temperature in the fixed effects.
  \item[c] Last period's actual values used as this period's forecast.
  \end{footnotesize}
\end{tablenotes}
\end{threeparttable} 
\end{center}
\vspace{1cm}

\doublespacing

\section*{References}
\bibliography{referencesFS}
